---
title: "Shallow Groundwater"
author: "Tiffany Le"
date: "2026-06-11"
output: pdf_document
---

```{r}
library(dplyr)
```

```{r}
water_year <- function(date) {
  y <- as.numeric(format(date, "%Y"))
  m <- as.numeric(format(date, "%m"))
  ifelse(m >= 10, y + 1, y)
}

season <- function(month) {
  if (month %in% c(12, 1, 2, 3)) return(1)
  else if (month %in% c(4, 5, 6)) return(2)
  else if (month %in% c(7, 8, 9)) return(3)
  else if (month %in% c(10, 11)) return(4)
}

plot_seasons <- function(df, x) {
  par(mfrow = c(2, 2))
  formula <- as.formula(paste("dtgw ~", x))
  for (s in 1:4) {
    sub <- df[df$season == s,]
    sub <- sub[is.finite(sub$dtgw) & is.finite(sub[[x]]),]
    if (nrow(sub) < 3) {
      plot.new()
      next
    }
    plot(formula, data = sub, main = season_labels[s])
  }
}
```

```{r}
season_labels <- c("CSR", "PMD", "MON", "PMR")

cameras <- data.frame(
  site = c(
    "Boquillas",
    "CharlestonMesquite",
    "Contention",
    "Fairbank",
    "Hereford",
    "Hunter",
    "Moson",
    "St.David"
  ),
  camera = c(8, 2, 6, 1, 5, 4, 3, 7)
)

filter <- c(1, 2, 6, 7)
```

```{r}
points <- read.csv("../Data/pts_for_Jawad.csv")[1:8,]
points$Name <- gsub(" Camera", "", points$Name)
points$Index <- seq_len(nrow(points))

pmdl <- do.call(rbind, lapply(points$Index, function(i) {
  file <- paste0("../Data/Jawad_ET/J", i, ".csv")
  et <- read.csv(file)
  et$site <- points$Name[i]
  et
}))

pmdl <- merge(pmdl, cameras, by = "site")
pmdl$date <- as.Date(pmdl$date)
pmdl$et <- pmdl$LE_model * 0.0353 # from ft to m
pmdl$wy <- water_year(pmdl$date)
pmdl$season <- sapply(as.integer(format(pmdl$date, "%m")), season)
pmdl <- pmdl[order(pmdl$camera, pmdl$date),]

et_seasonal <- pmdl %>%
  group_by(camera, wy, season) %>%
  summarise(et = sum(et, na.rm = TRUE), .groups = "drop")
```

```{r}
flow <- read.csv("../Data/USPPFlowMonitoring2006_2025.csv", stringsAsFactors = FALSE)
flow$Flow.Code[flow$Flow.Code == 2] <- 1
flow <- flow[flow$Flow.Code %in% c(0, 1),]
flow$Date <- as.Date(flow$Date)
flow <- merge(flow, cameras, by.x = "Site", by.y = "site")
flow$wy <- water_year(flow$Date)
flow$season <- sapply(as.integer(format(flow$Date, "%m")), season)

fp <- flow %>%
  group_by(camera, wy, season) %>%
  summarise(fp = mean(Flow.Code, na.rm = TRUE), .groups = "drop")
```

## WHIP DTGW Data

WHIP depth to groundwater data recommended by Fern, averaged across the three nearest wells per camera.

```{r}
mapping <- read.csv("../Data/GW/CamNearest3GWGage.csv", stringsAsFactors = FALSE)

dtgw <- read.csv("../Data/GW/ShallowGW2006_2025.csv", stringsAsFactors = FALSE) %>%
  filter(site_tp_cd == "GW", parameter_cd == "72019") %>%
  select(station_nm, lev_timestamp, lev_va) %>%
  inner_join(mapping, by = c("station_nm" = "station")) %>%
  transmute(
    camera,
    date = as.Date(lev_timestamp, format = "%m/%d/%Y"),
    dtgw = lev_va * 0.3048, # from ft to m
    wy = water_year(date),
    season = sapply(as.integer(format(date, "%m")), season)
  ) %>%
  arrange(camera, date)
```

```{r}
fp_seasonal <- fp %>%
  filter(camera %in% filter) %>%
  group_by(wy, season) %>%
  summarise(fp = mean(fp, na.rm = TRUE), .groups = "drop")

gw_seasonal <- dtgw %>%
  filter(camera %in% filter) %>%
  group_by(wy, season) %>%
  summarise(dtgw = mean(dtgw, na.rm = TRUE), .groups = "drop")

plot_seasons(inner_join(fp_seasonal, gw_seasonal, by = c("wy", "season")), "fp")
```

FP vs. GW by camera:

```{r}
gw_cam <- dtgw %>%
  group_by(camera, wy, season) %>%
  summarise(dtgw = mean(dtgw, na.rm = TRUE), .groups = "drop")

fp_gw <- inner_join(fp, gw_cam, by = c("camera", "wy", "season"))

for (cam in sort(unique(fp_gw$camera))) {
  plot_seasons(fp_gw[fp_gw$camera == cam,], "fp")
}
```

Seasonal ET vs. monsoon GW averaged across cameras 1, 2, 6, 7:

```{r}
gw_mon <- dtgw %>%
  filter(camera %in% filter, season == 3) %>%
  group_by(wy) %>%
  summarise(dtgw = mean(dtgw, na.rm = TRUE), .groups = "drop")

et_avg <- et_seasonal %>%
  filter(camera %in% filter) %>%
  group_by(wy, season) %>%
  summarise(et = mean(et, na.rm = TRUE), .groups = "drop")

plot_seasons(inner_join(et_avg, gw_mon, by = "wy"), "et")
```

Seasonal ET vs. monsoon GW by camera:

```{r}
gw_mon_cam <- dtgw %>%
  filter(season == 3) %>%
  group_by(camera, wy) %>%
  summarise(dtgw = mean(dtgw, na.rm = TRUE), .groups = "drop")

et_gw <- inner_join(et_seasonal, gw_mon_cam, by = c("camera", "wy"))

for (cam in sort(unique(et_gw$camera))) {
  plot_seasons(et_gw[et_gw$camera == cam,], "et")
}
```

## Continuous DTGW Data

USGS daily well-based depth to groundwater.

```{r}
dtgw_continuous <- read.csv("../Data/GW/C4GW.csv", stringsAsFactors = FALSE) %>%
  mutate(
    date = as.Date(date),
    camera = as.integer(camera),
    wy = water_year(date),
    season = sapply(as.integer(format(date, "%m")), season)
  )
```

FP vs. GW for C4:

```{r}
gw_cam <- dtgw_continuous %>%
  group_by(camera, wy, season) %>%
  summarise(dtgw = mean(dtgw, na.rm = TRUE), .groups = "drop")

fp_gw <- inner_join(fp, gw_cam, by = c("camera", "wy", "season"))

plot_seasons(fp_gw[fp_gw$camera == 4,], "fp")
```

Seasonal ET vs. monsoon GW for C4:

```{r}
gw_mon_cam <- dtgw_continuous %>%
  filter(season == 3) %>%
  group_by(camera, wy) %>%
  summarise(dtgw = mean(dtgw, na.rm = TRUE), .groups = "drop")

et_gw <- inner_join(et_seasonal, gw_mon_cam, by = c("camera", "wy"))

plot_seasons(et_gw[et_gw$camera == 4,], "et")
```

## Groundwater Level Delta Data

Height above the lowest recorded reading from the nearest USGS gauge to each camera with data following 2017.

```{r}
gw_delta <- read.csv("../Data/GW/CamSPRNCA_GW.csv", stringsAsFactors = FALSE) %>%
  rename(dtgw = dtgw_m) %>%
  mutate(
    date = as.Date(datetime),
    wy = water_year(date),
    season = sapply(as.integer(format(date, "%m")), season)
  ) %>%
  arrange(camera, date)
```

FP vs. GW averaged across cameras 1, 2, 6, 7:

```{r}
fp_seasonal <- fp %>%
  filter(camera %in% filter) %>%
  group_by(wy, season) %>%
  summarise(fp = mean(fp, na.rm = TRUE), .groups = "drop")

gw_seasonal <- gw_delta %>%
  filter(camera %in% filter) %>%
  group_by(wy, season) %>%
  summarise(dtgw = mean(dtgw, na.rm = TRUE), .groups = "drop")

plot_seasons(inner_join(fp_seasonal, gw_seasonal, by = c("wy", "season")), "fp")
```

FP vs. GW by camera:

```{r}
gw_cam <- gw_delta %>%
  group_by(camera, wy, season) %>%
  summarise(dtgw = mean(dtgw, na.rm = TRUE), .groups = "drop")

fp_gw <- inner_join(fp, gw_cam, by = c("camera", "wy", "season"))

for (cam in sort(unique(fp_gw$camera))) {
  plot_seasons(fp_gw[fp_gw$camera == cam,], "fp")
}
```

Seasonal ET vs. monsoon GW averaged across cameras 1, 2, 6, 7:

```{r}
gw_mon <- gw_delta %>%
  filter(camera %in% filter, season == 3) %>%
  group_by(wy) %>%
  summarise(dtgw = mean(dtgw, na.rm = TRUE), .groups = "drop")

et_avg <- et_seasonal %>%
  filter(camera %in% filter) %>%
  group_by(wy, season) %>%
  summarise(et = mean(et, na.rm = TRUE), .groups = "drop")

plot_seasons(inner_join(et_avg, gw_mon, by = "wy"), "et")
```

Seasonal ET vs. monsoon GW by camera:

```{r}
gw_mon_cam <- gw_delta %>%
  filter(season == 3) %>%
  group_by(camera, wy) %>%
  summarise(dtgw = mean(dtgw, na.rm = TRUE), .groups = "drop")

et_gw <- inner_join(et_seasonal, gw_mon_cam, by = c("camera", "wy"))

for (cam in sort(unique(et_gw$camera))) {
  plot_seasons(et_gw[et_gw$camera == cam,], "et")
}
```
